Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-05-07 manu-draft.qmd
gam summary plot main text added

GAMs to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass change was influenced by climate stressors. The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.

@@

![Predicted results from generalized linear models evaluating seagrass changes versus temperature or salinity predictors for the (a) EPC and (b) FIM datasets. Results are shown only for OTB, although the shape of the relationships do not change between bay segments. The predictors for (a) were the number of days when temperature was above the threshold and the number of days when salinity was below the threshold. The predictors for (b) were measured temperature and salinity at the bottom of the water column. Time periods were seagrass recovery prior to 2016 and seagrass decline after 2016. Shaded areas are 95% confidence intervals. Points are observed data that include aggregated effects of bay segment included in the models.](figs/sgmods.png){#fig-sgmod}

>>>>

GAMs to assess the potential effects of year, light attenuation, temperature and salinity on seagrass change provided modest evidence that seagrass change was influenced by climate stressors. The model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.

@@

![Predicted results from Generalized Additive Models evaluating seagrass changes versus year, temperature, and salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, n = 81, R.sq. 0.81, deviance explained 89%) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, n = 19, R.sq. 0.69, deviance explained 81%) datasets. Partial effects describe the modeled association between each predictor and seagrass change after accounting for the effects of the other predictors. The models also included smoothed interaction terms with year (not shown). See Tables S9 and S10 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmods.png){#fig-sgmod}
